---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# discounting: Temporal discounting models to understand affect dynamics

This project revolves around the comparison of several temporal discounting models when applied to affect, following up on the call for such a comparison mentioned in Vanhasbroeck et al. (2024). The code within this project combines the structure of an R package alongside scripts that call upon the functions used within that package, providing its own ecosystem in which functions evolve together with our research purposes. While in-principle the `discounting` package can be used by others, it is not necessarily meant to and may, therefore, feel user-unfriendly. 



## Installation

To install the package and use its functions, one can use the `remotes` package:

```
remotes::intall_github("ndpvh/discounting-affect")
```

One can then call the `library()` function to load the package's functions:

```{r}
library(discounting)
```



## Getting Started

Most of the package's functionality depends on the definition of an instance of a model, namely of the exponential discounting model ([`exponential()`](https://ndpvh.github.io/discounting-affect/reference/exponential.html); Samuelson, 1937, Rutledge et al., 2014), the quasi-hyperbolic discounting model ([`quasi_hyperbolic()`](https://ndpvh.github.io/discounting-affect/reference/quasi_hyperbolic.html); Laibson, 1997), or the double-exponential discounting model ([`double_exponential()`](https://ndpvh.github.io/discounting-affect/reference/double_exponential.html); van den Bos & McClure, 2013). For example, we can create a one-dimensional exponential discounting model as follows:

```{r}
my_model <- exponential(
  parameters = list(
      "alpha" = 1, 
      "beta" = as.matrix(1),
      "gamma" = as.matrix(0.5)
  ),
  covariance = as.matrix(1)
)
my_model
```

Simulating from this model can be achieved through the [`simulate()`](https://ndpvh.github.io/discounting-affect/reference/simulate.html) function:

```{r}
data <- simulate(
  my_model,
  X = runif(100, min = -1, max = 1)
)
data
```

To estimate parameters of a model based on observed data, we have to transform our dataset to an instance of the [`dataset`](https://ndpvh.github.io/discounting-affect/reference/dataset.html) class and can then call the function [`fit()`](https://ndpvh.github.io/discounting-affect/reference/fit.html) for a particular instance of the models denoted above:

```{r}
#| cache = TRUE
# Define the model you wish to fit
fit_model <- quasi_hyperbolic(d = 1, k = 1)

# Fit the model to the previously simulated data
result <- fit(
  fit_model,
  data,
  trace = FALSE
)
result$model 
```



## Analyses

The analysis workflow applies the exponential, quasi-hyperbolic, and double-exponential discounting models to four affect datasets (`VANHASBROECK_2021`, `VANHASBROECK_2022`, `VANHASBROECK_2024`, and `NIEMEIJER_2022`). The main aim is to compare how well the alternative discounting functions describe affective dynamics, while also examining the estimated time course of affect, the recoverability of the model parameters, and whether the fitted models reproduce important characteristics of the observed data.

All active analysis code is located in `analysis/`. The recommended entry point for reviewers is `analysis/00_run_analysis.R`, which runs the numbered scripts in their intended order and keeps each step in a fresh R session. From the repository root it can be run with `Rscript analysis/00_run_analysis.R`. By default, it uses `MODE = "existing_results"`: the generation stages (`01`--`04`) are skipped and the saved computational results are used to regenerate the downstream summaries and figures. Changing this to `MODE = "full_reproduction"` reruns the complete workflow, including data processing, model estimation, recovery, and the parametric bootstrap.

The numbered scripts correspond to the following stages:

* `01_process_data.R`: Converts the raw datasets into the participant-level `dataset` objects used for estimation.
* `02_estimate_models.R`: Fits all three discounting models separately for each participant and saves parameter estimates and fit statistics, including AIC and BIC.
* `03_run_recovery.R`: Performs the parameter-recovery study to evaluate whether known simulated parameter values can be recovered by the estimation procedure.
* `04_run_parametric_bootstrap.R`: Simulates data from the fitted models to test whether they can reproduce relevant empirical characteristics of the datasets.
* `05_model_comparison.R`: Compares the three models using participant-level AIC and BIC and summarizes how often each model provides the best fit.
* `06_forgetting_steps.R`: Translates the estimated temporal parameters into forgetting steps, defined as the number of time steps required for an effect to fall below the selected threshold.
* `07_nonparametric_bootstrap_negativity_bias.R`: Quantifies uncertainty in the model-comparison results and tests whether negative affect is forgotten more slowly than positive affect.
* `08_parameter_summary.R`: Summarizes the distributions of the estimated model parameters across participants and flags estimates concentrated near their parameter bounds.
* `09_summarize_recovery.R`: Converts the saved recovery objects into summary statistics and recovery figures without rerunning the recovery study.
* `10_summarize_parametric_bootstrap.R`: Summarizes the saved parametric bootstrap results, including model coverage of the empirical statistics.
* `11_visualization.R`: Produces the figures used to present the model, model-comparison, forgetting-step, and bootstrap results.

The shared files `_config.R` and `_helpers.R` define paths, dataset/model names, and common helper functions used by the numbered scripts; they are not separate analysis stages. Generated numerical results are stored under `analysis/results/`, with separate folders for estimation, model comparison, forgetting steps, parameter summaries, recovery, and bootstrap analyses. Generated figures are stored under `analysis/figures/`. The scripts themselves contain more detailed comments on dataset-specific preprocessing, model specification, and individual outputs.




## Further Reading

One can consult the package's vignettes for more information on the project, the [models](https://ndpvh.github.io/discounting-affect/articles/models.html), and the [package](https://ndpvh.github.io/discounting-affect/articles/getting_started.html).



# License

The work in this repository is shared under a dual license. The GNU GPL-3.0 license applies to the analysis code, the package code, and most of the datasets (VANHASBROECK_2021, VANHASBROECK_2022, VANHASBROECK_2024). The CC BY-NC 4.0 license applies to the other datasets (NIEMEIJER_2022), disallowing the commercial use of these data without reference to or contacting the original authors (koen.niemeijer@kuleuven.be).

`SPDX-License-Identifier: GPL-3.0 OR CC BY-NC 4.0`



# References 

Laibson, D. (1997). Golden eggs and hyperbolic discounting. _The Quarterly Journal of Economics, 112_(2), 443-477.

Samuelson, P. A. (1937). A note on measurement of utility. _The Review of Economic Studies, 4_(2), 155-161.

Rutledge, R. B., Skandali, N., Dayan, P., & Dolan, R. J. (2014). A computational and neural model of momentary subjective well-being. _Proceedings of the National Academy of Sciences of the United States of America, 111_(33), 12252–12257. doi: <a href = "https://doi.org/10.1073/pnas.1407535111">10.1073/pnas.1407535111</a>

van den Bos, W. & McClure, S. M. (2013). Towards a general model of temporal discounting. _Journal of the Experimental Analysis of Behavior, 99_, 58–73. doi: <a href = "https://doi.org/10.1002/jeab.6">10.1002/jeab.6</a>

Vanhasbroeck, N., Loossens, T., & Tuerlinckx, F. (2024). Two peas in a pod: Discounting models as a special case of the VARMAX. _Journal of Mathematical Psychology_, Article 102856. doi: <a href = "https://doi.org/10.1016/j.jmp.2024.102856">10.1016/j.jmp.2024.102856</a>
