---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# discounting: Temporal discounting models to understand affect dynamics

This project revolves around the comparison of several temporal discounting models when applied to affect, following up on the call for such a comparison mentioned in Vanhasbroeck et al. (2024). The code within this project combines the structure of an R package alongside scripts that call upon the functions used within that package, providing its own ecosystem in which functions evolve together with our research purposes. While in-principle the `discounting` package can be used by others, it is not necessarily meant to and may, therefore, feel user-unfriendly. 



## Installation

To install the package and use its functions, one can use the `remotes` package:

```
remotes::intall_github("ndpvh/discounting-affect")
```

One can then call the `library()` function to load the package's functions:

```{r}
library(discounting)
```



## Getting Started

Most of the package's functionality depends on the definition of an instance of a model, namely of the exponential discounting model ([`exponential()`](https://ndpvh.github.io/discounting-affect/reference/exponential.html); Samuelson, 1937, Rutledge et al., 2014), the quasi-hyperbolic discounting model ([`quasi_hyperbolic()`](https://ndpvh.github.io/discounting-affect/reference/quasi_hyperbolic.html); Laibson, 1997), or the double-exponential discounting model ([`double_exponential()`](https://ndpvh.github.io/discounting-affect/reference/double_exponential.html); van den Bos & McClure, 2013). For example, we can create a one-dimensional exponential discounting model as follows:

```{r}
my_model <- exponential(
  parameters = list(
      "alpha" = 1, 
      "beta" = as.matrix(1),
      "gamma" = as.matrix(0.5)
  ),
  covariance = as.matrix(1)
)
my_model
```

Simulating from this model can be achieved through the [`simulate()`](https://ndpvh.github.io/discounting-affect/reference/simulate.html) function:

```{r}
data <- simulate(
  my_model,
  X = runif(100, min = -1, max = 1)
)
data
```

To estimate parameters of a model based on observed data, we have to transform our dataset to an instance of the [`dataset`](https://ndpvh.github.io/discounting-affect/reference/dataset.html) class and can then call the function [`fit()`](https://ndpvh.github.io/discounting-affect/reference/fit.html) for a particular instance of the models denoted above:

```{r}
#| cache = TRUE
# Define the model you wish to fit
fit_model <- quasi_hyperbolic(d = 1, k = 1)

# Fit the model to the previously simulated data
result <- fit(
  fit_model,
  data,
  trace = FALSE
)
result$model 
```



## Analyses

For those who wish to reproduce or otherwise inspect our analyses, they can find the analysis scripts in the folder `scripts`. Currently, we provide the following analyses:

- `recovery.R`: Recovery study, providing a sanity check of being able to estimate the models provided by the package.



## Further Reading

One can consult the package's vignettes for more information on the project, the [models](https://ndpvh.github.io/discounting-affect/articles/models.html), and the [package](https://ndpvh.github.io/discounting-affect/articles/getting_started.html).



# References 

Laibson, D. (1997). Golden eggs and hyperbolic discounting. _The Quarterly Journal of Economics, 112_(2), 443-477.

Samuelson, P. A. (1937). A note on measurement of utility. _The Review of Economic Studies, 4_(2), 155-161.

Rutledge, R. B., Skandali, N., Dayan, P., & Dolan, R. J. (2014). A computational and neural model of momentary subjective well-being. _Proceedings of the National Academy of Sciences of the United States of America, 111_(33), 12252–12257. doi: <a href = "https://doi.org/10.1073/pnas.1407535111">10.1073/pnas.1407535111</a>

van den Bos, W. & McClure, S. M. (2013). Towards a general model of temporal discounting. _Journal of the Experimental Analysis of Behavior, 99_, 58–73. doi: <a href = "https://doi.org/10.1002/jeab.6">10.1002/jeab.6</a>

Vanhasbroeck, N., Loossens, T., & Tuerlinckx, F. (2024). Two peas in a pod: Discounting models as a special case of the VARMAX. _Journal of Mathematical Psychology_, Article 102856. doi: <a href = "https://doi.org/10.1016/j.jmp.2024.102856">10.1016/j.jmp.2024.102856</a>
